---
title: "R: Basic Operations"
output: html_notebook
---

# R: Basic Operations

## Comments

Comments in R are denoted by `#`

```{r}
# This is a comment
```

There is no built-in way to do multi-line comments.

## Mathematical Operations

Mathematical operations in R are handled using the standard 4 symbols `+, -, /, and *` . Operations are calculated using PEMDAS.

### Addition

```{r}

sum <- 5 + 3
print(sum)  

total <- 1 + 2 + 3 + 4
print(total)  

```

### Subtraction

```{r}
# Subtraction
difference <- 10 - 4
print(difference) 

result <- 20 - 5 - 3
print(result)  

```

### Multiplication

```{r}
# Multiplication
product <- 6 * 7
print(product)

result <- 2 * 3 * 4
print(result)

```

### Division

```{r}
# Division
quotient <- 20 / 4
print(quotient)  

result <- 100 / 5 / 2
print(result)  

```

## Variable Assignment

Variables in R are traditionally assigned using an arrow sign `<-` , made using a less than carat and a dash. It is valid to use `=` , but not recommended.

```{r}

a <- 5
b <- 10
c <- a + b
print(c)  

result <- sqrt(16)
print(result)  

numbers <- c(1, 2, 3, 4, 5)
print(numbers) 

data <- data.frame(name = c("John", "Jane"), age = c(30, 25))
print(data)

```

Variable names can not include and spaces or symbols other than the underscore `_` . They cannot begin with numbers, but can include numbers after the first character.

```{r}

total_score <- 95
average_temperature <- 22.5
student_name <- "John Doe"
data_set <- data.frame(a = 1:3, b = 4:6)

```

## Data Types

R has the following data types: numeric, integer, character, logical, and complex.

### Numeric

Decimal values in R are numerics. Numerics are the default data type for numbers.

```{r}
x <- 10.5
y <- 3.14
z <- 0.001

print(x)  
print(y)  
print(z)  


```

```{r}
a <- 5.5
b <- 2.2

sum <- a + b        
difference <- a - b  
product <- a * b     
quotient <- a / b   

print(sum)       
print(difference) 
print(product)    
print(quotient)   

```

Note: Numerics are not integers. You can confirm this with both a and b above using the `is.integer()` function.

```{r}

a <- 5.5

b <- 2L

print(is.numeric(a)) 

print(is.integer(a))  

print(is.numeric(b))  

print(is.integer(b))  

```

```{r}

a <- 5.5
b <- 2L


print(is.numeric(a))  
print(is.integer(a))  

print(is.numeric(b))  
print(is.integer(b)) 


c <- 7.89
d <- as.integer(c)
print(d)  
print(is.integer(d))  
 

```

### Integer

To create an integer in R we can use the `as.integer()` function.

```{r}
c <- 3.14
d <- as.integer(c)
print(d)  
print(is.integer(d))  
```

```{r}

a <- 5.5
b <- 2L

print(is.numeric(a)) 
print(is.integer(a))  

print(is.numeric(b))  
print(is.integer(b))  

c <- 7.89
d <- as.integer(c)
print(d) 
print(is.integer(d))  

```

For shorthand, we can also define an integer by writing the desired integer value, followed by a capital L.

```{r}
integer_var <- 5L

print(integer_var) 


print(is.integer(integer_var)) 

print(is.numeric(integer_var))  

```

```{r}
numeric_var <- 5

integer_var <- 5L

print(is.numeric(numeric_var)) 
print(is.integer(numeric_var))  

print(is.numeric(integer_var))  
print(is.integer(integer_var))  

```

# Example 2: Extract fixed-length substring

string \<- "Hello, World!"

# Extract substring starting from position 7 to position 11

substring \<- substr(string, start = 7, stop = 11) print(substring) \# Output: "World"

```{r}

str_invalid <- "abc123"


numeric_value <- as.numeric(str_invalid)
print(numeric_value)  
print(is.numeric(numeric_value))  


integer_value <- as.integer(str_invalid)
print(integer_value)  
print(is.integer(integer_value))  # Output: TRUE, but value is NA

```

We would use an integer data-type when we want to represent a whole integer, and don't want to accept any other variations of numbers. The numeric data-type is less strict, it allows us to switch between different number types like integers and floats dynamically.

### Character

A **character** type is used to represent strings in R. Characters can be denoted by writing within quotes.

```{r}

char_single <- 'Hello, World!'
print(char_single)  


char_double <- "Hello, R!"
print(char_double) 

```

#### Concatenating character objects

Character objects can be concatenated into a single variable using the `paste` function

```{r}
# Concatenating strings
concatenated <- paste("Hello", "World!")
print(concatenated)  

```

#### Converting into character objects

We can convert other objects into character objects using the `as.character()` function

```{r}

num <- 123
char <- as.character(num)
print(char)  

```

```{r}
# Note that class returns the data type of our object

logical_value <- TRUE
char <- as.character(logical_value)
print(char)  

```

#### Printing character types

You can print out hardcoded character objects with the `print` function

```{r}

print("Hello, World!")

```

You can print out formatted character objects with the `sprintf` function. You can use the following symbols to represent formatted parts of the code, each should be preceded by a `%` symbol:

-   d and i - Integer values

-   s - Character string

-   o - Octal

-   x and X - Hexadecimal

-   f - Double precision value (For fixed point representation, for example: 21.234)

-   e and E - Double precision value (For exponential representation, for example: 2.1234e+01)

```{r}

num <- 123
formatted <- sprintf("Integer value: %d", num)
print(formatted)  

```

```{r}

name <- "Alice"
age <- 30
formatted <- sprintf("Name: %s, Age: %d", name, age)
print(formatted)  

```

#### Return a substring

You can get a substring from a character object using the `substr` function

```{r}


string <- "Hello, World!"


substring <- substr(string, start = 7, stop = 11)
print(substring)  
 

```

#### Replacing characters in a string

You can replace characters in a string using the `sub` function

```{r}

original_string <- "Hello, World!"
modified_string <- sub("o", "0", original_string)

print(modified_string)

```

### Logical

A logical value is commonly referred to as a boolean value. It is the result of a comparison between two variables. It is represented as `TRUE` or `FALSE.`

```{r}
x <- 5
y <- 10


result <- x < y

print(result)  

```

```{r}

is_true <- TRUE
is_false <- FALSE

print(is_true)  
print(is_false)  

```

#### Logical Operators

The standard logical operations are:

-   `&` - And

-   `|` - Or

-   `!` - Not

```{r}

x <- TRUE
y <- FALSE
print(x & y)  

```

```{r}

x <- TRUE
y <- FALSE
print(x | y)  

```

```{r}

x <- TRUE
print(!x)  

```

### Complex

A complex object in R is created using the symbol `i` . The use of the i symbol is in reference to imaginary numbers.

```{r}

z <- complex(real = 3, imaginary = 4)
print(z)  


```

```{r}

z <- as.complex("3+4i")
print(z)  

```

## Comparison Operators

You can use the following operators to compare two variables and return a logical object:

-   `<` Less than

-   `>` Greater than

-   `<=` Less than or equal to

-   `>=` Greater than or equal to

-   `==` Is equal to

-   `!=` Is NOT equal to

```{r}
x <- 5
y <- 10
print(x < y)  

```

```{r}
x <- 5
y <- 10
print(x > y)  

```

```{r}
x <- 5
y <- 5
print(x <= y)  

```

```{r}
x <- 5
y <- 5
print(x >= y)  

```

```{r}
x <- 5
y <- 5
print(x == y) 

```

```{r}
x <- 5
y <- 10
print(x != y)  

```

## Homework

1.  Create two strings with your first and last name, respectively. Then concatenate them, and print the new value as "My name is \<your newly created concatenated variable will go here\>"

    ```{r}
    first_name <- "John"
    last_name <- "Doe"

    full_name <- paste(first_name, last_name)


    print(paste("My name is", full_name))

    ```
